Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm

### 2024-09-23 notebooks/04-response-to-reviewers.qmd

> We thank the reviewer for this interesting interrogation. We computed a correlation between the two effects, and it is not significant ($\rho$ = 0.05, 95% CI \[-0.11, 0.21\], *p* = 0.57). Thus, the two effects indeed appear to be separable to some extent. We added this analysis in the Methods on p.18, l.1 and the Results on p.27, l.1.

Minor suggestions for the figures: Could the authors include some transitional probabilities in figure 2, so it is immediately clear which stimuli follow with which probabilities. The figure now seems to suggest that there is no preceding cue in the black gabor conditions, which I know is not the case from reading the methods.

Further, I think the data figures would be improved if they included individuals data points (violin or raincloud plots perhaps). Also, the plotted effect is a within subject effect, so within standard errors of the mean would be a more intuitive presentation of the results, rather than standard errors of the mean.

> We thank the reviewer for these suggestions. Figure 2 has been adjusted on p.12. For the response times Figures 5 and 6, the choice of representing the marginal means for the groups only was based on the important gap between individual mean response times (approximately ranging from 500 to 1000ms) and the effect size of interest (\~30ms). We added global figures with individual data points and "zoomed" versions to highlight the effect of interest for each task. This resulted in Figures 5 on p.21, 6 on p.22, 8 on p.25 and 9 on p.26.

@@

When conducting Pearson correlations across groups, you need to control for group as in the presence of large mean differences in groups, a correlation can be driven by group differences, rather than a true linear relationship between the variables. In line with this, the claim about a continuum in the discussion should be softened.

> We thank the reviewer for spotting this oversight. We also later realised that the linearity assumption was not justified in our problem, especially considering the ordinal nature of the data and the underlying group structure mentioned. To relax this linearity assumption, we instead conducted Spearman correlations (i.e., on ranked data). Correlations with the OSIQ-O and SUIS remained significant, the correlation with the VVIQ became significant by a slight margin, and the conclusions were nuanced appropriately. Ranked correlations indeed allowed us to discuss the *monotonicity* (preservation of order) of the link, but not the linearity along a continuum, as has been pointed out here. The analytic process was corrected in the Methods on p.18, l.1-11, the Results have been modified on pp.26-27, and the discussion of these results has been adjusted on p.30, l.15-21.

@@

Since a claim is being made about the existence of a null-effects in the aphantasia groups, it might be useful to report bayes factors in favour of the null-hypothesis.

> We thank the reviewer for suggesting Bayesian analyses. We fitted Bayesian models independently for aphantasic and control groups to assess the evidence in favour or against the existence of a congruence effect in these "sub-models". They confirmed strong evidence in favour of a null congruence effect in the aphantasia group in the implicit task ($BF_{01} = 14.12$) and explicit task ($BF_{01} = 12.09$), and extreme evidence in favour of a congruence effect in the control group in the implicit task ($BF_{01} < .001$) and explicit task ($BF_{01} < .001$). The description of the Bayesian modelling procedure can be found on p.17 l.1-13, and the results are mixed with the reports of model contrasts (and highlighted) throughout the Results.

@@

Could the authors please discuss the limitations of their response mapping procedure. The responses aren't randomized (horizontal is always H and vice versa), which means participants could in theory learn the relationship between a cue and a button-press, rather than the effect of a cue on visual processing speed. How could this change the interpretation of the findings?

> Indeed, H was consistently Horizontal and V was Vertical. However, we would like to emphasize that there were as many congruent trials as incongruent ones, which, in our view, implies that participants could not reliably predict the correct response based on the cue alone. If they had responded solely based on the key they associated with the cue, only 50% of the responses would have been correct, which likely would have prevented the observation of a congruence effect. In light of this observation, we do not believe that learning the response key associated with the cue was a relevant or adopted strategy by the participants. Rather, we argue that the effect we measured was indeed related to the influence of a cue on visual processing speed, in line with our interpretation. This methodological consideration was added in the procedure details on p.11 l.8-14.

@@

Were there any differences in accuracy? Even if not, it might still be useful to report.

> We thank the reviewer for this suggestion. We initially removed errors as there were very few and they are usually not analysed in RT tasks, however we agree that this precision could be of interest for the reader to have a comprehensive picture of the results. Consequently, we analysed accuracy using logistic regression models to assess potential differences between the groups, before subsequently removing errors for RT analyses. Details of the analytic methods have been added on p.15, l.1-10 to describe the models, and results have been added on pp.19-20. There were no accuracy differences between the groups.

@@

# Reviewer 2

In this manuscript, the authors report on an experiment where they have developed a new implicit priming paradigm to test the hypothesis that aphantasics have unconscious mental imagery. Their results support that aphantasics do not have unconscious mental imagery, as they do not demonstrate a priming effect on the task. This is an interesting and well-designed paradigm which gets at a phenomenon which is not often tested in aphantasia research, and as such I think the manuscript would make an important contribution to the literature. However, I believe there are points which warrant substantial revisions in both the analysis conducted and the presentation of the work, and for this reason I would recommend acceptance following major revisions to the manuscript. Below I list points for the authors to address divided into major and minor issues, and I look forward to hearing the authors' responses to the concerns I raise.

"They found no differences in accuracy between the groups, but slower RTs and lower confidence in the aphantasics' answers and argued that this result was consistent with aphantasics having the visual images required to complete the task, albeit without conscious access to them." (p2) The authors here claim that visual images are required to complete the task, but it is not clear that this is in fact the case as alternative hypotheses aren't ruled out. It would be possible to encode information in a different format (e.g. linguistically) and still perform this task. Further, it is unclear why slower RT from aphantasic participants would imply that they still have visual images. Could the authors please clarify the connection between slower RT and lack of mental imagery.

> We thank the reviewer for this comment on the study of @liuProbingUnimaginableImpact2023. We acknowledge that any interpretation remains delicate and should be approached with caution when a task can be completed by encoding information in an alternative format (e.g., linguistically). It is particularly important to be mindful of this possibility in the tasks being proposed. In our own study, we made a concerted effort to account for this possibility by including a condition in which the Gabors had no color, thus preventing participants from employing the linguistic strategy of learning that all red Gabors had a specific orientation and responding solely based on their color. As for the slower reaction times (RTs) observed in aphantasic individuals, this also remains difficult to interpret, and caution is warranted. Rather than attributing the slower RTs to the use of mental imagery, it is possible that alternative strategies, which take more time to implement, could explain the slower RTs. Liu and Bartolomeo themselves link slower RTs to lower confidence levels and suggest that this link reflects an inability to analyze their own mental processes, or in other words, to "know how they did it." However, we agree that this interpretation should be approached with caution.

> We reformulated to try and make this more explicit on p.2, l.17-20: "They found no differences in accuracy between the groups, but slower RTs and lower confidence in the aphantasics' answers, and argued that this result was consistent with aphantasics having the visual images required to succeed at the task, albeit without the knowledge of "how they did it"."

@@

> We thank the reviewer for this thorough recommendation and for spotting this oversight. The intuition here was correct: our sample comprises "total" aphantasics (VVIQ = 16, N = 50), "hypophantasics" [terminology of @reederNonvisualSpatialStrategies2024, VVIQ between 17 and 32, N = 39], and "controls" (VVIQ between 32 and 74) but only two hyperphantasics (VVIQ $\geq$ 75).

> Thus, we conducted a new analysis that could answer several questions raised here, by comparing these "finer-grained" groups of total aphantasics, hypophantasics, and controls, leaving out the two hyperphantasics. These analyses resulted in an interesting trend specific to the hypophantasic group, with a trend implicit effect but no explicit effect. We amended the introduction on p.6, l.12-23, the Methods on p.17, l.14-23, the Results on pp.23-26, and these new pattern are discussed on p.29, l.7-10. The conclusion on p.35 has also been modified accordingly.

@@

"We argue that this novel paradigm provides a strong base to develop implicit objective behavioural assessments of visual imagery, thereby opening promising avenues for a better objective characterization of aphantasia." (p25) What do authors mean by "better" here? Do they mean that it's better than binocular rivalry, galvanic skin response, automatic pupil dilation? If so, why? Is it more reliable? Please elaborate on what makes it better, and what it makes it better relative to.

> Studies using binocular rivalry, galvanic skin response, and automatic pupil dilation have greatly inspired us and made valuable contributions to the objective characterization of aphantasia. With our study, we aimed to provide an additional and original contribution, which, alongside the work of other teams, would contribute toward a better objective characterization of aphantasia. However, we wish to highlight the strengths and unique features of our paradigm, namely that it relies on implicit rather than explicit mental imagery, is easy to implement (requiring no expensive or complex equipment, just a computer), and can be conducted online, allowing for the recruitment of a wider sample. In our view, these strengths make our paradigm a significant contribution.

> The sentence has been modified to better reflect our thought on p.30, l.11-14: "We propose that this novel online paradigm provides a foundation to develop implicit, objective behavioural assessments of visual imagery using a minimal setup, thereby opening new avenues for a large-scale, objective characterization of aphantasia."

@@

It is not clear why including the non-coloured condition makes it unlikely that a propositional (rather than imagery based) strategy could have been used by participants with aphantasia (p28). Please clarify the reasoning here.

> Our phrasing likely led to the assumption that the non-colored condition made the use of a propositional strategy unlikely, and it requires clarification. Our point is rather that this strategy would only be advantageous in the colored condition: as soon as the color of the target appears, participants using this strategy respond with the key corresponding to the orientation associated with the color of the Gabor, an association they have memorized verbally (e.g., "red means horizontal"). However, when the non-colored target appears, participants can no longer rely on the color of the target to use the propositional strategy and are thus forced to process the orientation of the lines. Since the colored and non-colored trials were presented randomly, sticking to the propositional strategy could be hard, thus our hypothesis that it encouraged participants to abandon it. The misleading wording in our sentence was likely to be "our tasks had the advantage of *preventing* the use of these type of strategies", as our tasks do not strictly "prevent" anything. We nuanced it on p.33 l.2-3: "our tasks have the advantage of making the use of this type of strategies more difficult to adhere to, thanks to a condition in which the targets are not coloured."

@@

"Curiously, the analysis of questionnaire data yielded a significant difference between aphantasics and controls in reported spatial imagery, assessed by the spatial scale of the OSIQ, contrasting with previous studies on aphantasia that used this questionnaire and found no between-group differences (Bainbridge et al., 2021; Dawes et al., 2020; Keogh & Pearson, 2018)." (p29) The authors should take into account alternative explanations. For example, it could be explained by the different uses of cut-off points in aphantasia. The differing results here could potentially be explained by differences in how aphantasia groups are demarcated in different articles. The authors used VVIQ $\leq$ 16 to denote aphantasia, whereas the other studies use other cut-off points (see Blomkvist and Marks, 2023 and Blomkvist 2022 for discussions).

> The issue of the diversity of cut-off points chosen to define aphantasia across the literature is indeed pervasive and should be examined with care for any effects reported. However, the results reported here were those of the standard group defined by VVIQ $\leq$ 32, so this group did not depart much from those defined by the various studies cited. Thus, the statistical result found in the present study remains singular and puzzling, leading us towards the exploratory hypothesis we formulate later in the paragraph on p.33, l.10-23.

@@

> We thank the reviewer for this suggestion. We added a mention of Blomkvist's (2022) views to this paragraph on p.3, l.11-14 "On the other hand, several studies that sought to develop objective measures assessing visual imagery found consistent behavioural and physiological differences between aphantasics and controls on imagery tasks, thus challenging Nanay's (2020) hypothesis of unconscious mental imagery in aphantasia (see also Blomkvist, 2022). Aphantasics have been shown to have a reduced skin conductance response \[...\]"

> As mentioned earlier, the terms used throughout the article will be harmonised and the unconscious mental imagery at stake here was better defined on p.2, l.10-13.

The author cites the Jacobs et al. (2018) study as potential evidence for unconscious imagery in aphantasia, but ignore the results from Keogh and Pearson (2019) which speak against this (p2). This is moreover both discussed in Nanay (2020) and Blomkvist (2023) and would be relevant to mention here.

> The results speaking against this interpretation of Jacobs' experiment and the idea of unconscious imagery (Keogh & Pearson, 2018, cited also by Blomkvist, 2022) are detailed in the next paragraph. The structure we chose was to expose arguments brought forward by teams supporting the "unconscious imagery" hypothesis, then the opposite view. We apologize if this structure was unclear at the time of reading. We added additional details on the counter-arguments in the subsequent paragraph to makes this more clear on p.4, l.10-17.

@@

Final scoring is reported for the SUIS but not the VVIQ (p8). Amend for consistency.

> The VVIQ score range (16-80) was added on p.8, l.11.

"This result was also present when adopting a more conservative definition of aphantasia by analysing groups restricted to aphantasics scoring at floor VVIQ and controls with high VVIQ scores." (p24) It is misleading to say that a VVIQ of 42 is high. Most scores above 42 will be in the average range.

> This discussion was amended on p.29, l.7-10 to reflect the results of the new analyses on finer-grained VVIQ groups mentioned earlier that replaced this.

@@

"This result supports the hypothesis that aphantasia may be associated with a reduction or absence of mental image generation, even unconscious, as opposed to a lack of conscious access to mental images."(p24) "may be associated with" is too vague. Be clearer about the relationship.

> This sentence has been rephrased on p.29, l.9-10: "This result supports the hypothesis that aphantasics have difficulty generating both conscious and unconscious mental images, rather than only conscious ones. "

@@

"The explicit task with instructions to produce mental imagery was inspired by the binocular rivalry paradigm developed by Keogh & Pearson (2018) and managed to replicate their pattern of results showing an absence of priming in aphantasia, thus validating the effectiveness of priming tasks to evidence conscious mental imagery differences." (p24) It would be relevant here to point out that the priming effect from Keogh and Pearson (2018) is not the same as the priming effect in the current study, as the current study was not in fact assessing binocular rivalry by presenting different stimuli to the two eyes at the same time.

> We thank the reviewer for spotting this oversight. This precision was added on p.29, l.14-18: "The explicit task with instructions to produce mental imagery was inspired by the binocular rivalry paradigm developed by Keogh (2018) and produced a similar pattern of results showing an absence of priming in aphantasia with a different task, thus validating the effectiveness of priming tasks to evidence conscious mental imagery differences."

@@

"It is interesting to observe that our paradigm correlates with rather ecological forms of visual imagery." (p25) Do the authors mean to imply that the VVIQ items are not ecological? Most of them seem to be, comprising e.g., of imagining a person you know well or a shopfront. Please elaborate on this claim to clarify whether the VVIQ contains non-ecological items and why these should be taken to be non-ecological.

> Our wording was also inaccurate here. By "ecological", we meant to designate the broader scope of the SUIS and OSIQ, which also probe the use of different forms of imagery in daily life, whereas the VVIQ is focused on the immediate experience upon filling the questionnaire.

> However, based on a suggestion by another reviewer, we corrected our method for the correlational analyses and conducted Spearman (i.e., ranked) correlations instead of Pearson correlations to account for the ordinal nature of questionnaire responses and the inability to make the claim of a linear relationship between continuous questionnaires scores and congruence effects. The results of these new analyses showed that the correlations with the VVIQ were in the same range as those of the OSIQ and SUIS, but that they might be driven by group differences. As such, this argument about the nature of the VVIQ was not relevant anymore. The entire correlation results part has been corrected accordingly and this discussion paragraph changed entirely on p.30, l.15-21.

@@

"However, this raises the question of whether our study assessed "mental imagery", as it is commonly defined as a conscious and voluntary experience." (p26) Please provide citations for this claim. In fact, many definitions do not have a take on whether this has to be voluntary and conscious and instead claim the opposite (see for example, Pearson (2019), Nanay (2020)).

> We thank the reviewer for this comment. The sentence here was indeed inaccurate, as the problem we eventually discuss is whether the concept of "unconscious imagery" is relevant. This introductory sentence was revised on p.31, l.6-9.

The authors start section 4.2 by discussing whether their study is really targeting unconscious mental imagery and then go on to discuss claims made by Muraki et al. regarding sensorimotor stimulation (pp26-27). I'm not sure this is the most relevant material to draw on. Nanay's conception of unconscious mental imagery seems more appropriate. The final sentence in this section is also somewhat difficult to parse, please consider rephrasing.

> We thank the reviewer for this suggestion. Indeed, Nanay's (2020) views were very important in the field of aphantasia to define the concept of unconscious mental imagery which is at the heart of the present study. We added a discussion of these views in the same paragraph on p.31, l.10-15.

@@

"This type of strategy could even have been detrimental, which could explain slower response times in the uncoloured condition than in the coloured condition." (p28) What do authors mean by "detrimental" here, it seems like a strange choice of word? Consider rewording.

> The wording is indeed inaccurate here. We used the word in the meaning of "counter-productive", and replaced it accordingly on p.33, l.6.

@@

"Large differences in spatial imagery ability also existed within the aphantasic group, hinting that this finding could be specifically tied to our sample and, more generally, that there could exist various sub-types in aphantasia characterized by their variable reliance on different forms of mental representations (e.g., spatial, verbal, kinaesthetic)." (p29) This seems like a rather odd suggestion. It is commonplace to suggest that there are different kinds of aphantasia based on different sensory modalities (auditory, visual, olfactory, gustatory, tactile), as well as other systems (affective, motor). But something like "verbal aphantasia" seems like a contradiction in terms as aphantasia explicitly has to do with mental imagery (a form of mental representation) and a verbal representations are per definition not imagistic.

> We thank the reviewer for this comment. The word "verbal" was replaced by "auditory" on p.33, l.20, which is more relevant as a form of aphantasia.

@@

"Furthermore, it has not yet been proven that aphantasia is a pathological disorder Blomkvist & Marks (2023)." (p29) This point could be put a bit more delicately, as it seems to imply that researchers are trying to prove this and that we should expect that it will be proven. It would also be relevant to cite Monzel et al. (2023) here as this is the study which investigates the question, which Blomkvist and Marks then discuss.

> As we acknowledged in response to a previous comment, our tone was too rough in some places and lacked nuance. This was adjusted throughout the discussion, and indeed here on p.34, l.17: " Furthermore, it has been shown that aphantasia does not meet the criteria of a pathological disorder (Monzel et al., 2023; Blomkvist & Marks, 2023)."

### 2024-09-23 notebooks/04-response-to-reviewers.qmd

"In sum, our findings provided evidence suggesting that aphantasia does not only result from lacking metacognition but might reflect an actual alteration (whether a reduction or an absence) of both conscious and unconscious mental imagery." (p29) This conclusion is too broad as authors only tested VVIQ $\leq$ 16, and the parenthesis mention reduction in mental imagery (which I take to imply 17-32) but the authors only investigated a sample of participants with VVIQ $\leq$ 16.

> In line with previous comments, this final sentence on p.35, l.11-16 was adjusted to better reflect the results and nuance the findings: "In sum, our findings provide evidence suggesting that "total" aphantasia (defined by a VVIQ score of 16) may not solely result from impaired metacognition but could indicate an underlying reduction in both conscious and unconscious mental imagery. Additionally, a trend in the results indicated that hypophantasia (i.e., reduced imagery, VVIQ scores between 17 and 32) may involve the presence of unconscious imagery, albeit without the capacity for voluntary imagery generation."

### 2024-09-23 notebooks/03-manuscript-revised.qmd
Finished?

Some researchers doubt that mental imagery could be completely absent, and have hypothesized that aphantasics who report having no conscious imagery at all might still have *unconscious* mental imagery [@nanay_unconscious_2020]. As this phenomenon is difficult to identify, little is known about unconscious mental imagery in aphantasia. A recent study by @liuProbingUnimaginableImpact2023 used a behavioural task to assess visual imagery in various domains (the French Perception-Imagination Battery) in aphantasics and controls. The task consisted of mentally comparing pairs of items based on various features (e.g. "beaver" - "fox": which is the *longest*?). They found no differences in accuracy between the groups, but slower RTs and lower confidence in the aphantasics' answers, and argued that this result was consistent with aphantasics having the visual images required [to succeed at the task, albeit without the knowledge of "how they did it"]{bg-colour="#FFFF00"}. Similarly, in a task that supposedly required visual imagery to verify whether a target dot was inside the boundaries of a previously presented geometric shape, @jacobsVisualWorkingMemory2018 found no difference in accuracy between an aphantasic participant and controls. As proposed by @nanay_unconscious_2020, this result could be interpreted as reflecting an unconscious comparison of a visual image with the stimuli perceived by the aphantasic participant. [Although on a N = 1 case study, this result may still suggest]{bg-colour="#FFFF00"} the existence of this unconscious form of imagery in aphantasia. However, due to the potential conscious processes at play in these behavioural tasks (with or without sensory imagery), neither study could completely rule out the hypothesis of aphantasics using strategies other than visual imagery (e.g. semantic processing, spatial imagery) to solve the tasks. This possibility, often raised in discussions about aphantasia [[e.g., @jacobsVisualWorkingMemory2018; @knightMemoryImageryNo2022; @liuProbingUnimaginableImpact2023; @monzelImagineYouWill2021]{bg-colour="#FFFF00"}], prevents from firmly concluding from these results regarding the existence of unconscious mental images.

@@

Participants began with the implicit priming task. This task began with an associative phase (@fig-protocol-asso) in which a fixation cross (500 ms) was followed by one of the two coloured Gabor, and participants had to indicate the colour of the Gabor by pressing the corresponding key (either R for Red or B for Blue), without time constraint. Each Gabor was presented 50 times. After the association phase, participants started the implicit priming test phase (@fig-protocol-implicit) in which a fixation cross (500 ms) was followed by a cue (150 ms) that was either a red or a blue circle, and then depending on the condition by either one of the two coloured Gabor or by one of the uncoloured Gabor, as a target. Participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand. In the congruent condition, the cue has the same colour as the target Gabor (coloured condition) or primed the Gabor with the same line orientation as the black target Gabor (uncoloured condition), contrary to the non-congruent condition. There were 16 trials per congruence and colour pairs, amounting to 64 trials total for the task.

@@

A recurrent challenge when studying aphantasia is that many tasks that are often considered to require visual imagery could also be solved by other non-visual strategies, such as the use of internal or covert verbalization @monzelMemoryDeficitsAphantasics2022. In that vein, it could be argued that our tasks could be completed using a verbal or propositional strategy based on the colour of the stimuli. For instance, in the association phase, participants may have used the propositional strategy of forming the abstract or semantic representation of an association between colour and line orientation of the Gabors (e.g., by internally repeating that "blue patches have vertical lines"). During the tasks, participants could then have used such a representation when a prime was presented, which may have facilitated their response when the target's colour was congruent with the colour of the prime. However, our tasks have the decisive advantage of preventing the use of this type of strategy with a condition in which the targets are not coloured. As no semantic representations had been previously associated with uncoloured targets, a colour-based strategy was inappropriate for these stimuli. This type of strategy could even have been [counter-productive]{bg-colour="#FFFF00"}, which could explain slower response times in the uncoloured condition than in the coloured condition. Thus, the persistence of the interaction between group and congruence independently of the colour conditions supports the idea that our results can not be explained purely by the reliance on propositional strategies.

Curiously, the analysis of questionnaire data yielded a significant difference between aphantasics and controls in reported spatial imagery, assessed by the spatial scale of the OSIQ, contrasting with previous studies on aphantasia that used this questionnaire and found no between-group differences [@dawesCognitiveProfileMultisensory2020; @keoghBlindMindNo2018; @bainbridgeQuantifyingAphantasiaDrawing2021]. While this difference is hard to interpret on its own, it is worth noting that although significant, the difference in spatial imagery between the groups is much less pronounced than the difference observed on visual imagery scales. Large differences in spatial imagery ability also existed within the aphantasic group, hinting that this finding could be specifically tied to our sample and, more generally, that there could exist various sub-types in aphantasia characterized by their variable reliance on different forms of mental representations (e.g., spatial, verbal, kinaesthetic). Nevertheless, it should be mentioned that our paradigm relied partially on spatial judgements (namely, orientation properties of items) that might be impacted by spatial imagery abilities. This hypothesis may also apply to a growing number of studies on aphantasia that used paradigms consisting in the detection of orientation changes of Gabor patches, thought to assess visual imagery or visual working memory [e.g., @changBehavioralPerformanceCortical2023; @keoghAttentionDrivenPhantom2020; @keoghBlindMindNo2018; @keoghVisualWorkingMemory2021; @knightMemoryImageryNo2022; @slinnVividnessVisualImagery2023 inter alia]. While our results did exhibit differing patterns between groups, future studies should still pay particular attention to this spatial imagery factor. On the one hand, this would require a more systematic investigation of spatial imagery, which has proved to have distinct characteristics from visual imagery in aphantasia [see @blazhenkovaTwoEyesBlind2019; @palermoCongenitalLackExtraordinary2022]. On the other hand, this entails that care should be taken to design tasks that target visual imagery as specifically as possible, manipulating item properties such as colour or shape while accounting for semantic and spatial properties of items [see e.g., @liuProbingUnimaginableImpact2023].

>>>>

Some researchers doubt that mental imagery could be completely absent, and have hypothesized that aphantasics who report having no conscious imagery at all might still have *unconscious* mental imagery [@nanay_unconscious_2020]. As this phenomenon is difficult to identify, little is known about unconscious mental imagery in aphantasia. A recent study by @liuProbingUnimaginableImpact2023 used a behavioural task to assess visual imagery in various domains (the French Perception-Imagination Battery) in aphantasics and controls. The task consisted of mentally comparing pairs of items based on various features (e.g. "beaver" - "fox": which is the *longest*?). They found no differences in accuracy between the groups, but slower RTs and lower confidence in the aphantasics' answers, and argued that this result was consistent with aphantasics having the visual images required [to succeed at the task, albeit without the knowledge of "how they did it"]{bg-colour="#FFFF00"}. Similarly, in a task that supposedly required visual imagery to verify whether a target dot was inside the boundaries of a previously presented geometric shape, @jacobsVisualWorkingMemory2018 found no difference in accuracy between an aphantasic participant and controls. As proposed by @nanay_unconscious_2020, this result could be interpreted as reflecting an unconscious comparison of a visual image with the stimuli perceived by the aphantasic participant. [Although on a N = 1 case study, this result may still suggest]{bg-colour="#FFFF00"} the existence of this unconscious form of imagery in aphantasia. However, due to the potential conscious processes at play in these behavioural tasks (with or without sensory imagery), neither study could completely rule out the hypothesis of aphantasics using strategies other than visual imagery (e.g. semantic processing, spatial imagery) to solve the tasks. This possibility, often raised in discussions about aphantasia [[e.g., @jacobsVisualWorkingMemory2018; @knightMemoryImageryNo2022; @liuProbingUnimaginableImpact2023; @monzelImagineYouWill2021]]{bg-colour="#FFFF00"}, prevents from firmly concluding from these results regarding the existence of unconscious mental images.

@@

Participants began with the implicit priming task. This task began with an associative phase (@fig-protocol-asso) in which a fixation cross (500 ms) was followed by one of the two coloured Gabor, and participants had to indicate the colour of the Gabor by pressing the corresponding key (either R for Red or B for Blue), without time constraint. Each Gabor was presented 50 times. After the association phase, participants started the implicit priming test phase (@fig-protocol-implicit) in which a fixation cross (500 ms) was followed by a cue (150 ms) that was either a red or a blue circle, and then depending on the condition by either one of the two coloured Gabor or by one of the uncoloured Gabor, as a target. Participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand. In the congruent condition, the cue has the same colour as the target Gabor (coloured condition) or primed the Gabor with the same line orientation as the black target Gabor (uncoloured condition), contrary to the non-congruent condition. There were 16 trials per congruence and colour pairs, amounting to 64 trials total for the task. [Response mapping was not randomised due to the structure of the task: as there were as many congruent trials as incongruent ones, participants could not reliably predict the correct response based on the cue alone and a cue-response key association. Should they respond solely based on the key they associated with a cue, responses would be close to random, which could be easily spotted in the data. Consequently, learning a response key associated with a cue was not relevant and randomisation was not deemed necessary.]{bg-colour="#FFFF00"}

@@

A recurrent challenge when studying aphantasia is that many tasks that are often considered to require visual imagery could also be solved by other non-visual strategies, such as the use of internal or covert verbalization @monzelMemoryDeficitsAphantasics2022. In that vein, it could be argued that our tasks could be completed using a verbal or propositional strategy based on the colour of the stimuli. For instance, in the association phase, participants may have used the propositional strategy of forming the abstract or semantic representation of an association between colour and line orientation of the Gabors (e.g., by internally repeating that "blue patches have vertical lines"). During the tasks, participants could then have used such a representation when a prime was presented, which may have facilitated their response when the target's colour was congruent with the colour of the prime. However, [our tasks have the advantage of making the use of this type of strategies more difficult to adhere to, thanks to a condition in which the targets are not coloured.]{bg-colour="#FFFF00"} As no semantic representations had been previously associated with uncoloured targets, a colour-based strategy was inappropriate for these stimuli. This type of strategy could even have been [counter-productive]{bg-colour="#FFFF00"}, which could explain slower response times in the uncoloured condition than in the coloured condition. Thus, the persistence of the interaction between group and congruence independently of the colour conditions supports the idea that our results can not be explained purely by the reliance on propositional strategies.

Curiously, the analysis of questionnaire data yielded a significant difference between aphantasics and controls in reported spatial imagery, assessed by the spatial scale of the OSIQ, contrasting with previous studies on aphantasia that used this questionnaire and found no between-group differences [@dawesCognitiveProfileMultisensory2020; @keoghBlindMindNo2018; @bainbridgeQuantifyingAphantasiaDrawing2021]. While this difference is hard to interpret on its own, it is worth noting that although significant, the difference in spatial imagery between the groups is much less pronounced than the difference observed on visual imagery scales. Large differences in spatial imagery ability also existed within the aphantasic group, hinting that this finding could be specifically tied to our sample and, more generally, that there could exist various sub-types in aphantasia characterized by their variable reliance on different forms of mental representations (e.g., spatial, [auditory]{bg-colour="#FFFF00"}, kinaesthetic). Nevertheless, it should be mentioned that our paradigm relied partially on spatial judgements (namely, orientation properties of items) that might be impacted by spatial imagery abilities. This hypothesis may also apply to a growing number of studies on aphantasia that used paradigms consisting in the detection of orientation changes of Gabor patches, thought to assess visual imagery or visual working memory [e.g., @changBehavioralPerformanceCortical2023; @keoghAttentionDrivenPhantom2020; @keoghBlindMindNo2018; @keoghVisualWorkingMemory2021; @knightMemoryImageryNo2022; @slinnVividnessVisualImagery2023 inter alia]. While our results did exhibit differing patterns between groups, future studies should still pay particular attention to this spatial imagery factor. On the one hand, this would require a more systematic investigation of spatial imagery, which has proved to have distinct characteristics from visual imagery in aphantasia [see @blazhenkovaTwoEyesBlind2019; @palermoCongenitalLackExtraordinary2022]. On the other hand, this entails that care should be taken to design tasks that target visual imagery as specifically as possible, manipulating item properties such as colour or shape while accounting for semantic and spatial properties of items [see e.g., @liuProbingUnimaginableImpact2023].
